---
title       : dplyr
subtitle    : A gamechanger for data manipulation in R
author      : Kevin Ferris
job         : Montana State University
framework   : io2012        # 
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
assets:
  css:
    - "assets/css/styles.css"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
library(knitr)
library(plyr)
opts_chunk$set(echo = T, 
               message = F, 
               warning = F, 
               cache = T, 
               tidy = F, 
               comment = "#>", 
               fig.align = "center", 
               fig.width = 8, 
               fig.height = 5, 
               digits = 3, 
               size = "small")
options(show.signif.stars = F, 
        digits = 2)
set.seed(99)
```



## dplyr

A new R package written by Hadley Wickham and Romain Francois (Wickham and Francois, 2014)

It provides a "flexible grammar of data manipulation"

> * Identifies the most important tasks in data manipulation and makes them easy to use
> * Provides blazing fast performance by writing key code in C++
> * Uses the same interface to manipulate data frames, data tables, and databases

---
## Common functions

1. `tbl_df`: for printing data.frames nicely
2. `select`: for choosing columns in a dataset
3. `filter`: for choosing rows in a dataset
4. `arrange`: for reordering rows in a dataset
5. `mutate`: for adding new columns to the dataset
6. `group_by`: for creating a grouped dataset
7. `summarise`: for summarizing a dataset

---
## Baseball data for an example

Who hit the most home runs for one team since 1920?

```{r lahman}
library(Lahman)
head(Batting)
```


---
## 1. Print data.frames with `tbl_df()`

Prints nicely and prevents you from accidentally printing the whole dataset.

```{r tbl_df}
library(dplyr)
batting_df <- tbl_df(Batting)
batting_df
```

---
## 2. Select columns with `select()`

Subset columns by name

```{r select}
batting_s <- select(batting_df, playerID, yearID, teamID, G, H, X2B, X3B, HR)
batting_s
```

---
## 3. Filter rows with `filter()`

Chooses rows that match your criteria criteria

```{r filter}
batting_f <- filter(batting_s, yearID > 2000, HR > 10)
batting_f
```

---
## 4. Arrange rows with `arrange()`

Reorders rows

```{r arrange}
batting_a <- arrange(batting_f, yearID, desc(HR))
batting_a
```

---
## 5. Add new columns with `mutate()`

Adds new columns

```{r mutate}
batting_m <- mutate(batting_a, X1B = H - X2B - X3B - HR)
batting_m
```

---
## Steps so far

1. Turn data.frame into a tbl with `tbl_df`
2. Select columns with `select`
3. Select rows with `filter`
4. Arrange rows with `arrange`
5. Add new columns with `mutate`

---
## Nested, Unreadable Code

```{r nested-code}
mutate(
  arrange(
    filter(
      select(
        tbl_df(
          Batting), 
        playerID, yearID, teamID, G, H, X2B, X3B, HR), 
      yearID > 2000, HR > 10), 
    yearID, desc(HR)), 
  X1B = H - X2B - X3B - HR)
```

---
## Chaining

`dplyr` provides the `%>%` operator.  

```{r chaining}
x <- 1:3
prod(cumsum(x))
x %>% 
  cumsum() %>% 
  prod()
```

> * This executes one command, **THEN** executes a second command.

> * This makes your code much more readable.

---
## Chaining Example

```{r chain-ex}
Batting %>%                                                 # takes the Batting data.frame, THEN
  tbl_df() %>%                                              # turns it into a tbl_df, THEN
  select(playerID, yearID, teamID, G, H, X2B, X3B, HR) %>%  # selects the columns, THEN
  filter(yearID > 2000, HR > 10) %>%                        # filters the rows, THEN
  arrange(yearID, desc(HR)) %>%                             # reorders the rows, THEN
  mutate(X1B = H - X2B - X3B - HR)                          # adds columns
```

---
## 6) Grouping with `group_by()`

Created a grouped tbl where operations are performed "by group"

Not really useful on its own, so let's combine it with...

--- 
## 7) Summarizing with `summarise()`

Summarize each group in a tbl

```{r group-summarise}
batting_m %>%            # take the batting_m tbl_df, THEN
  group_by(yearID) %>%   # groups it by year, THEN
  summarise(sum(HR)) %>% # sums HR for each yearID, THEN
  head()                 # prints the first six years
```

---
## `dplyr` is much, much, much, faster

```{r comparison}
library(microbenchmark)
microbenchmark(dplyr = batting_m %>% group_by(yearID) %>% summarise(sum(HR)), 
               plyr = ddply(batting_m, .(yearID), summarise, sum(HR)), 
               agg = aggregate(batting_m$HR, list(year = batting_m$yearID), sum), 
               tapp = tapply(batting_m$HR, list(batting_m$yearID), sum), 
               times = 500L)
```

---
## `dplyr` is even faster than `tapply` when there are many groups

```{r comparison2}
microbenchmark(dplyr = batting_m %>% group_by(yearID, playerID) %>% summarise(sum(HR)),
               tapp = tapply(batting_m$HR, list(batting_m$yearID, batting_m$playerID), sum), 
               times = 500L)
```


---
## So who hit the most home runs for one team since 1920?

```{r home-runs}
result <- Batting %>% 
  tbl_df() %>% 
  filter(yearID >= 1920) %>% 
  group_by(playerID, teamID) %>% 
  summarise(HR = sum(HR), 
            years = paste(min(yearID), max(yearID), sep = ":")) %>% 
  ungroup() %>% 
  arrange(desc(HR))
head(result)
```
<!--
---
## Joining with the `join()` verbs

`dplyr` implements SQL style joins with `inner_join()`, `left_join()`, `semi_join()`, and `anti_join()`

```{r join}
inner_join(batting_m, Master)
```
-->

---
## Other cool `dplyr` features

* `dplyr` implements SQL style joins

```{r join-2}
inner_join(result, Master %>% select(playerID, nameLast)) %>% head()
```

* can connect to SQL databases, run SQL code outside of R, then pull the results into R
* `dplyr` uses the same commands whether the data are stored in a database or loaded into R

---
## Other cool `dplyr` features

* other people are actively providing enhancements to `dplyr`
   * `tidyr` for reshaping data
   * `assertr` for making assertions about the data
   * `magrittr` and `pipeR` for chaining
   * `Rcpp` for speeding things up

---
## Other Resources


1. [dplyr cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
2. `dplyr`'s vignettes `browseVignettes(package = "dplyr")`
3. Introductions: 
    * [5 minute intro](http://practicalrvideos.blogspot.com/2014/08/dplyr-gamechanger-for-data-manipulation.html)
    * [Wickham's tutorial](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) (open the dplyr-tutorial.pdf document)
    * [Vivek Patil's thorough tutorial](http://patilv.com/dplyr/)
4. [Migrating from `plyr`](https://github.com/jimhester/plyrTodplyr)
5. [Some general thoughts on the dplyr API](http://datascience.la/dplyr-some-more-reflections/)
6. [More advanced example and plotting](http://timelyportfolio.github.io/rCharts_factor_analytics/factors_with_new_R.html)
7. [`dplyr's GitHub repo](https://github.com/hadley/dplyr)



